\chapter{User manual}
In this section it will be explained how to use the software developed for the purpose of the experiments. First section describes the parameters 

\section{Input parameters}
System is accessed through a single class to which command line parameters are passed in the form:
\begin{center}
[-argumentType   argumentValue]*
\end{center}

\textbf{Class to which the arguments should be passed is  called $Experiment$.  }

All arguments are optional. All arguments have their default values. Class that parses arguments is called $Arguments$.
An overview of all input arguments is given in the table below.

\begin{table}[h!]
\begin{footnotesize}
\begin{tabular}{| l|c|c|c|c|}
\hline
   index  &  name & data type & default value & used for  \\
\hline\hline
	1  &  lowercase & String & "n"  & preprocessing \\ \hline
	2  &  stopWordsRemoval & String & "n"  & preprocessing \\ \hline
	3  &  stemming & String & "n"  & preprocessing \\ \hline
	4  &  mergeLexicalVariants & String & "n"  & preprocessing \\ \hline
	5  &  inputFilePath & String & pdt1\_cleaned  & getting data \\ \hline
	6  &  numberOfSentencesInLuceneDoc & Integer & 1  & document size \\ \hline
	7  &  numberOfWordsInDocument & Integer & -1  & document size \\ \hline
	8  &  matrixType & Integer & 0  & \small{\{0,1,2\}=\{TFIDF,PMI,RI\}} \\  \hline
	9  &  upBoarderForNumberOfMeanings & Integer & 6  & evaluation \\ \hline
	10  &  evaluationContextWindowSize & Integer & 3  & evaluation \\ \hline
\hline
\end{tabular}
\caption{Overview of system's input arguments}
\end{footnotesize}
\end{table}

If someone wishes to perform WSD on another dataset, they should specify the path to the file with the
$-inputFilePath$ argument. However, it should be noted that occurrences of polysemous words should
be annotated so that different meanings could be differentiated by the system. Content of the file
 passed to the system is split into three separate files: train, testDev, testFinal. TestFinal contains test
test sentences. 


\section{Logs}
For every experiment a log is made in the log folder, containing all relevant statistics related to the 
experiment. Log name is constructed from abbreviations of all values of experiment's phases, for instance: $PMI\_STOP+MERGE\_1s\_7w\_6m$ means that it is a log of PMI model, where in preprocessing
 filtering out closed class words and merging lexical variants was performed, document size in term--doc matrix is 1 sentence, 7+7 evaluation window context was used to construct context vectors, and only words with 6 meanings were investigated.